Case-based Arabic Morphological Analyzer for Arabic Text Mining

### 2010-12-20 tex/rp.tex
final version

Research in text mining started in the mid 1980s when Swanson 
realized that slicing and combining seemingly unrelated medical 
articles led to the discovery of new 
hypotheses~\cite{JNi06}.
Till present, most of the interest in text mining comes 
from biological sciences.
Lately we saw applications of text mining in the advertisement, 
political campaigning, and other businesses.
Text mining involves information retrieval (IR), natural language 
processing (NLP), information extraction (IE), 
and data mining~\cite{ASr09}.

@@

\begin{itemize}\itemsep0pt
\item CATARM will build a formal Arabic computational model
using efficient and configurable data structures such as 
finite state machines (FSM). 
\item CATARM will provide the user with a visual query 
language that makes it easy to enter relational queries. 
The query language will also allow the framework to extract Arabic
specific features from the query to simplify the linguistic 
computational model. 
\item CATARM will use statistical models and techniques
to infer language independent features from the text documents
that will further simplify the computational model. 
\item During the project, we will augment current Arabic digital 
dictionaries to resolve entity semantics and targeted 
qualifications such as adjectives.
\item The framework will allow the integration of library and 
custom computational components built on top of primitive
CATARM components. 
The components will be the linguistic computational model,
statistical, set arithmetic, logic,
and graph analysis techniques, and  text algorithms.
\item We will use CATARM to solve several realistic case studies.
We will look at the hadith literature and build a system using
CATARM to relate the hadith literature to the biographies of the
narrators and extract an authenticity graph for the narrators.
We will look at investigation and interrogation security reports
and build a CATARM system that looks for relations between suspects
in the reports. 
We will look at clinical reports and relate them to pharmaceutical 
records and find relations between sicknesses and prescribed medication. 
Finally, we will use CATARM to provide content-based parental control
for Arabic text documents. We will look for 
relations between documents requested by a minor and
a set of undesired Arabic documents. 
If the relation is strong, we will flag access to
the documents.

@@

CATARM will discover relations between Arabic text documents that 
are hard to discover by humans and will have direct live and
cost saving benefits.% in cases similar to the above mentioned 
%case studies. 
CATARM will qualitatively enhance research in the cultural and 
literature fields. 
It may also help investors analyze Arabic news and data from Arab 
markets prior to making important decisions.
CATARM may be used to flag spammers 
posting biased reviews in Arabic web forums.

@@

The framework now comprises a state of the art
case-based morphological analyzer, ATSarf, 
and computational components that allow the user to
specify his queries as one or more FSM
controlling the morphological analyzer.
We explain this approach in details in 
Section~\ref{s:design:method}.
ATSarf along with other computational components, 
extract the chains of narrators from a set of narrations
with above 95 percent accuracy and in less than two minutes.

@@

The directed acyclic graph (DAG) 
in Figure~\ref{f:narrators} shows a partial order relation (POR) 
between narrators.
We automatically extracted the POR from three books of hadith 
selected arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}
\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }.
The nodes in boxes are the \RL{matn} of the hadith, 
and the other nodes are the narrators.
Given a set of chains of narrators similar to 
Figure~\ref{f:exhadith} we formed the DAG using graph algorithms 
that merged equal names in the chains. 
This partial order graph is instrumental to automate
localizing narrators in biography documents and
segmenting biography documents.

@@

Once the biographies are analyzed, one can annotate
the POR with qualifiers of the narrators that reflect
their authenticity. 
One can also annotate the POR with the locations and 
the time they lived in. 
Then one can compute a time and location overlap
check to discover inconsistent narrations.
We can also perform several interesting checks using 
this POR on its own such as checking for the effect of
one narrator on the hadith literature.

@@

The specific aims of this project are:
\begin{itemize}
\item Build the CATARM framework while solving four interesting Arabic 
text minining case studies
  \begin{itemize}
  \item Hadith literature case study
  \item Security case study
  \item Network content case study
  \item Health case study
  \end{itemize}
\item Build and enhance a case-based morphological analyzer
\item Build an Arabic linguistic computational model (LCM)
\item Build entity and relation extractors
\item Build a query language 
  \begin{itemize}
  \item Build optimization modules that use the query to reduce the LCM
  \item Build a query solver that compiles all the components and run them 
  \end{itemize}
\item Build language independent LCM optimization modules
\item Add statistical, graph, and text manipulation algorithms
\item Build a visualization tool to showcase the results 
\end{itemize}

In the rest of this section, we will introduce Arabic text mining in 
details, then we will discuss the case studies in 
Sections~\ref{s:design:lit}--~\ref{s:design:med}. 
We will discuss the morphological analyzer in Section~\ref{s:design:ma}, 
the query language in Section~\ref{s:design:query}, and the linguistic
computational model in Section~\ref{s:design:lcm}.
In Section~\ref{s:design:method}, we will discuss our methodology
to build the components, optimize and reduce the LCM, build the
entity and relation extractors, solve the query, and visualize the 
results. 
Section~\ref{s:design:plan} will details the work plan and
justify the proposed budget.

@@

To do the above, 
we need tools that automatically differentiate between 
names, verbs, adjectives and other structures.
This is not a simple task with the Arabic language.
For example, Arabic grammar allows name-based sentences 
or sentences with no verbs \noTrNoVocRL{^gml ismiyT} .
Another unique feature is the possibility to have names in Arabic 
that express action such as noun verbs 
\noTrNoVocRL{ism f`l, f-a`l $\ldots$}.

In addition to these structural differences,
there are features in the Arabic language that need special 
treatment such as derivatives and stems 
\noTrNoVocRL{alm^staq-at w al^g_dwr}.
Simple primitive tasks such as comparing two strings need 
special attention in Arabic since each Arabic letter has 
four different forms depending on its position in the word; 
isolated form, start, middle and end of a word.
Arabic diacritics should also be ignored most of the time 
when comparing two strings.
We also need to build dictionary tools that associate words 
and phrases based on their meanings and their context.
Often time,
and depending on the application and the user of the tool,
we need to allow the user to assign meanings and semantics 
for patterns in the text.

@@

The computational components can be direct CATARM components
or custom user components built from other CATARM components.
We feed the hadith books to a user Narrator Chain Extractor
component that we will discuss later in 
Section~\ref{s:design:method}. We  obtain a set of chains
of narrators that we feed to a morphological analyzer (MA) 
component controlled by a merger controller. 
The merger controller will look at the morphological analysis
of the narrator names and will decide to merge two narrators in 
one graph node in case they were morphologically equal. 
It will also give feedback to the MA component to stop the
analysis when a it reaches a satisfactory result.
We consider this graph as a partial order relation between 
narrators where a narrator precedes another narrator if 
he narrated from him.

@@

@@

The diagram in Figure~\ref{f:security} describes our 
high level vision for what the solution of the 
security case study
will look like in the CATARM framework. 
The inputs are two separate sets of documents. 
The first set includes crime scene investigation reports, 
and the second set includes a 
interrogation reports with crime suspects as well as 
a database of previous interrogation reports. 
The CATARM system also takes as input a database of geographical
locations and a database of personal records. 
The aim is to check for inconsistencies in the interrogation
reports and to identify possible connections between the suspects
and the crime scene.

A morphological analyzer with date, place, and name extraction controller extracts
names, dates and places from the investigation reports. 
We pass the dates to a sorting component to produce $G_D$ 
a partial order graph between the dates. 
We look up the names in the personal records database and find out
$G_N$ a family dependency graph between the names. 
We pass the places to a mapper that builds another $G_G$,
a graph relating the places using a geographical database.

We feed $G_D, G_F, G_G,$ and $G_R$   
to a consistency checker.
The consistency checker considers the three dimensions 
$N$ names, $G$ locations, and $D$ dates and projects $G_R$
over each one of them.
For each dimension $k$, it computes a local measure 
$\alpha_k$ of how much $G_R|_k$ simulates $G_k$. 
It flags the components in the graph where the accumulated 
measure $\alpha= \Sigma_{k\in\{D,N,G\}} \alpha_k$ is
below a threshold $\theta$ as inconsistent.

@@

Parental controls help parents allow their kids to benefit 
from the Internet 
without fearing the danger of violent and obscene material. 
These controls partially rely on processing the content to detect 
and rate the target material. 
Control mechanisms for Arabic traffic as currently mostly manual. 
In this case study we will detect the bad content. 
Given a set of undesired Internet content 
and a set of requests by Internet users, 
CATARM can discover whether there is a relation between 
the requested documents and the undesired content and flag the 
requested document with strong enough relation. 
The undesired content is available online and is in the 
order of thousands of documents. 
We will look at Arabic Internet traffic five times larger to
validate our case study.

@@

Clinical reports contain diagnosis information, observations about 
patient progress, prescriptions, and patient response to prescribed 
medications. 
In countries, like Lebanon, where medical doctors graduate and 
specialize 
in several distinct practices around the world it is interesting to 
check for common practices and emerging trends to gain insight
about local and regional health and diagnosis features. 
It is also interesting to relate 
pharmaceutical 
records to clinical reports to check on 
market consistency and on self prescribed medicine 
practices.
Such a study can uncover serious tendencies in uncontrolled
pharmaceutical markets.
It can uncover what kind of sicknesses and what kind of medicines
are used sensitive medicines are used without 
appropriate medical supervision.

@@

Morphological analyzers for the Arabic language
such as Buckwalter~\cite{Buckwalter:02},
Beesley~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} take an Arabic word as input
and produce several valid morphological solutions for the word. 
Each solution is a sequence of morphemes that form the word.
Each morpheme is associated with several tags such as 
morphological, part of speech, grammar, and semantic tags. 
These analyzers are essential in several text analysis tools such 
as spell checkers as well as NLP frameworks~\cite{Col09}.
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy at the expense of efficiency.

@@

Case-based analysis can guide the morphological analysis
to prune lots of the solutions and to bias the order of considered
decisions. 
For example, in the hadith literature case study presented above, 
amongst other things, we are interested in checking whether a 
word is a name.  
We can bias the order of considered decisions to consider the 
morphemes that can be names first. 
If one of them matches, we can ignore the rest of the solutions. 
In case a prefix is decided to be of a morphological category
that does not connect with a name, then we can just ignore
the analysis of the rest of the word. 
This is more efficient than generating all possible partitions
of the word, checking for morphological correctness of each 
partition, and then inspecting the part of speech tags of each
correct solution. 
We desire to build a morphological analyzer that allows a 
case-based controller to intervene at every single decision
without much overhead.

Toutanova in~\cite{Tou09} suggested that using 
a joint model to do lemmatization and part of speech (POS
tagging performs better
than the direct and the pipelined approach. 
We take the first argument further to suggest a full joint NLP task 
model via using the NLP query to simplify and render the analysis
more accurate at the low level morphological analysis.

@@

We propose the CATARM query language with a user friendly
visual representation similar to that in Figures~\ref{f:literature} and
~\ref{f:security}. 
The basic elements in the query language will be entities, relations
and tags similar to those reported by a morphological analyzer. 
The query language will also have connectives such as set operations, 
relational operations, 
scalar and Boolean arithmetic operations, 
statistical computations, graph algorithms, and text algorithms
form terms and predicates that specify the desired computation
by the user. 
We will specify the CATARM query language in a formal grammar.

@@

We think that we can discover such features by building
a linguistic computational model (LCM) of the Arabic language
based on its formal grammar and formal morphology
as defined in several Arabic linguistic 
sources~\cite{Sha73,Abd00,Abd001}.
We will build our morphological analyzer on top 
of this LCM addition to the current existing
concatenative-based analysis techniques.

Our LCM will differ from ElixirFM~\cite{Otakar:07}
in that it will follow a relational approach rather 
than a functional morphological approach. 
That is the solutions will be expressed in a directed acyclic graph 
(DAG)
representing a hierarchy of relations between the morphemes 
rather than morphological trees mapping the morphemes to the 
solutions.
We think this provides a more succinct
representation for the solutions.
Also interesting relations 
between morphemes can be computed statically before
the input is presented to the analyzer.
Morphological trees are traversals of these DAGs, are expensive 
to exhaustively compute and store statically, 
and are expensive to compute dynamically. 
Our intuition is that Arabic allows relational structures
such as noun sentences and complex hierarchical entities.

Let $L_q$ be the query language grammar, 
$L_A$ be the grammar that expresses the Arabic linguistic
computational model, and let $q$ be a user query.
We compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We will explore methods that use $\rho_q$ to simplify 
$L_A$ before we start the analysis. 
We think the possible simplifications that we can perform
here are Language specific as they will reflect the 
correspondence between the formal query language definition $L_q$, 
the usage patterns of $L_q$  by Arabic users, 
and $L_A$ the formal LCM for Arabic.

@@

{\bf Case-based morphological analyzer.~~}
We experimented with several morphological analyzers.
We think that using finite state machines for morphological analysis
can be implemented in a more efficient way compared to other 
approaches. 
For example the concatenative rules used in Buckwalter~\cite{Tim04}
can be encoded in three FSM with linear execution time using 
the trie~\cite{Aoe:89} data structure.

The use of FSMs also suits our case-based approach. 
The controller can be a FSM machine on its own and its decision
will be another input into the internal morphological FSM structures. 
This also suits our choice of enriching the concatenative analysis
with a formal linguistic model that can also be encoded into FSMs. 
FSM can also be easily integrated with hidden Markov models (HMM)
extracted from the sets of documents the MA will consider.

@@

In ~\cite{Dao09}, Daoud observes that the pipeline model,
where morphology is performed first and syntactic and other
processes follow, is not adequate for Arabic. 
He suggests a synchronized syntactic and morphological
approach.  
We take the argument further to 
allow a tighter integration between the several pipeline
levels up to the NLP query in question by providing a 
case-based morphological analyzer, a linguistic computational
model of Arabic, and a query language that can all be expressed
as FSMs. 
The work~\cite{AlSham08} reports a similar result where
the automated addition of syntax knowledge improved stemming
in terms of accuracy and efficiency.

@@

{\bf Simplify LCM based on the query.~~}
Given $L_q$ the query language, 
$L_A$ the Arabic linguistic
computational model, and  $q$ a user query
we compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We plan to inspect methods to over-approximate 
the morphemes and the transitions in $L_A$ 
that can be satisfied by morphemes and transitions
in $\rho_q$. 
This can simplify the number of states we need to 
look at under $L_A$.

@@

We will consider the use of HMM~\cite{JanHMM06}
models built on the fly while processing the documents.
A high level HMM model of the document builds an HMM state
model where the HMM states are morphemes or partial 
morphemes and the HMM edges represent transitions between
the HMM states. 
The HMM edges are labeled with weights that reflect the
We will look at building a correspondence between 
the HMM model and the LCM and then simplify the 
LCM where the aggregated probabilities happen to be
below a certain threshold. 
The work in~\cite{JanHMM06} uses a similar approach
to correspond English text to tags corresponding to the 
Unified Medical Language System (UMLS).

@@

{\bf Entity and relational extractor.~~}
We will build a set of entity and relational extractors. 
The entities can be proper names, dates, locations,
and qualification adjectives. 
We will explore the use of pairing techniques to establish 
relations between the extracted entities.
The relations will form graphs as in the case studies presented
so far. The nodes represent the entities and the previously 
extracted relations and the edges represent the newly discovered 
relations.

@@

We will need to augment the tags in the lexicons of the 
morphological analyzer to allow for better entity and relational
extractors. 
We plan to make use of existing annotated tree banks and of
modern Arabic dictionaries to do that.

{\bf Statistical components.~~}
We will make use and build statistical components that can be used
in CATARM. 
We will need statistical components to rank solutions, compute 
probable decisions, estimating good thresholds.

@@

The solver passes the query to a front end that builds
a parse graph of the query. 
It uses the parse graph to simplify the LCM and the MA
as discussed before. 
Then it traverses the resulting parse graph. 
When the solver meets an atomic element of the query, 
it will call the morphological analyzer to compute 
morphological matches to the atomic element.
When the solver meets an operation of the query, 
it will call the corresponding computational component 
and passes to it the solutions computed so far from children
nodes. 
The solver will use existing off the shelf constraint solvers 
and set theory satisfiability solvers to address the queries. 
The solver may make use of existing 
propositional~\cite{MiniSAT04} or
higher order logic~\cite{Z308} solvers if needed. 
These solvers can solve satisfiability questions.

@@

We will follow a relaxed  eXtreme Programming 
methodology in the process of building CATARM.
This means that we will define refined incremental items of 
work that allows for more frequent deliverables. 
This also means that the code base will always be subject
to maintenance and refactoring to enhance reuse and 
increase productivity. 
While an item is delivered, it will still be subject to work,
enhancements, and maintenance. 
We selected eXtreme Programming as this fits well in the 
context of research where the clients are at close 
contact with the developers.
It accomodates the high dependency amongst the 
modules of CATARM. 
The pair programming practice in eXtreme Programming 
works well in the context of academy and research to keep
expertise in the team. 
As senior students graduate, their junior partners in the pair
will keep the knowledge in.

>>>>

Research in text mining started in the mid 1980s when Swanson 
realized that slicing and combining seemingly unrelated medical 
articles led to the discovery of new 
hypotheses~\cite{JNi06}.
Till present, most of the interest in text mining comes 
from biological sciences.
Lately we saw applications of text mining in the advertisement, 
political campaigning, and other businesses.
%Text mining involves information retrieval (IR), natural language 
%processing (NLP), information extraction (IE), 
%and data mining~\cite{ASr09}.

@@

\begin{itemize}\itemsep0pt
\item CATARM will build a formal Arabic computational model
using efficient and configurable data structures such as 
finite state machines (FSM). 
\item CATARM will provide the user with a visual query 
language that makes it easy to enter relational queries. 
The query language will also allow the framework to extract Arabic
specific features from the query to simplify the linguistic 
computational model. 
\item CATARM will use statistical models and techniques
to infer language independent features from the text documents
that will further simplify the computational model. 
\item During the project, we will augment current Arabic digital 
dictionaries to resolve entity semantics and targeted 
qualifications such as adjectives.
\item The framework will allow the integration of library and 
custom computational components built on top of primitive
CATARM components. 
The components will be the linguistic computational model,
statistical, set arithmetic, logic,
graph analysis techniques, and  text algorithms.
\item We will use CATARM to solve several realistic case studies.
We will look at the hadith literature and build a system using
CATARM to relate the hadith literature to the biographies of the
narrators and extract an authenticity graph for the narrators.
We will look at investigation and interrogation security reports
and build a CATARM system that looks for relations between suspects
in the reports. 
We will look at clinical reports and relate them to pharmaceutical 
records and find relations between sicknesses and prescribed medication. 
Finally, we will use CATARM to provide content-based parental control
for Arabic text documents. We will look for 
relations between documents requested by a minor and
a set of undesired Arabic documents. 
If the relation is strong, we will flag access to
the documents.

@@

CATARM will discover relations between Arabic text documents that 
are hard to discover by humans and will have direct live and
cost saving benefits. % in cases similar to the above mentioned 
%case studies. 
CATARM will qualitatively enhance research in the cultural and 
literature fields. 
It may also help investors analyze Arabic news and data from Arab 
markets prior to making important decisions.
CATARM may be used to flag spammers 
posting biased reviews in Arabic web forums.

@@

The framework now comprises a state of the art
case-based morphological analyzer, ATSarf, 
and computational components that allow the user to
specify his queries as one or more FSM
controlling the morphological analyzer.
We explain this approach in details in 
Section~\ref{s:design:method}.
ATSarf along with other computational components, 
extract the chains of narrators from a set of narrations
with above 95 percent accuracy and in less than 10 seconds.

@@

The directed acyclic graph (DAG) 
in Figure~\ref{f:narrators} shows a partial order relation (POR) 
between narrators.
We automatically extracted the POR from three books of hadith 
selected arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}
\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }.
The nodes in boxes are the \RL{matn} of the hadith, 
and the other nodes are the narrators.
Given a set of chains of narrators similar to 
Figure~\ref{f:exhadith} we formed the DAG using graph algorithms 
that merged equal names in the chains. 
This partial order graph is instrumental to automate
localization of narrators in biography documents and
segmenting biography documents.

@@

Once the biographies are analyzed, one can annotate
the POR with qualifiers of the narrators that reflect
their authenticity. 
One can also annotate the POR with the locations and 
the time the narratores lived in. 
Then one can compute a time and location overlap
check to discover inconsistent narrations.
We can also perform several interesting checks using 
this POR on its own such as checking for the effect of
one narrator on the hadith literature.

@@

The specific aims of this project are:
\begin{itemize}
\item Build the novel CATARM framework while solving four interesting Arabic 
text minining case studies
  \begin{itemize}
  \item Hadith literature case study
  \item Security case study
  \item Network content case study
  \item Health case study
  \end{itemize}
\item Build and enhance a novel case-based morphological analyzer
\item Build a novel partial Arabic linguistic computational model (LCM)
\item Build entity and relation extractors
\item Build a query language 
  \begin{itemize}
  \item Build novel simplification modules that use the query to reduce the LCM
  \item Build a query solver that compiles all the components, runs them, and ranks them
  \end{itemize}
\item Build novel language independent LCM simplification modules
\item Add statistical, graph, and text manipulation algorithms
\item Build a visualization tool to showcase the results 
\end{itemize}

In the rest of this section, we will introduce Arabic text mining in 
details, then we will discuss the case studies in 
Sections~\ref{s:design:lit}--~\ref{s:design:med}. 
We will discuss the morphological analyzer in Section~\ref{s:design:ma}, 
the query language in Section~\ref{s:design:query}, and the linguistic
computational model in Section~\ref{s:design:lcm}.
In Section~\ref{s:design:method}, we will discuss our methodology
to build the components, optimize and reduce the LCM, build the
entity and relation extractors, solve the query, and visualize the 
results. 
Section~\ref{s:design:plan} will detail the work plan and
justify the proposed budget.

@@

To do the above, 
we need tools that automatically differentiate between 
names, verbs, adjectives and other elements.
This is not a simple task with the Arabic language.
For example, Arabic grammar allows name-based sentences 
or sentences with no verbs \noTrNoVocRL{^gml ismiyT} .
Another unique feature is the possibility to have names in Arabic 
that express action such as noun verbs 
\noTrNoVocRL{ism f`l, f-a`l $\ldots$}.

In addition to these structural differences,
and to difficulties mentioned in Section~\ref{s:background},
there are features in the Arabic language that need special 
treatment such as derivatives and stems 
\noTrNoVocRL{alm^staq-at w al^g_dwr}.
Simple primitive tasks such as comparing two strings need 
special attention in Arabic since each Arabic letter has 
four different forms depending on its position in the word; 
isolated form, start, middle and end of a word.
Arabic diacritics should also be ignored most of the time 
when comparing two strings.
We also need to build dictionary tools that associate words 
and phrases based on their meanings and their context.
Often time,
and depending on the application and the user of the tool,
we need to allow the user to assign meanings and semantics 
for patterns in the text.

@@

The computational components can be direct CATARM components
or custom user components built from other CATARM components.
We feed the hadith books to a user Narrator Chain Extractor
component that we will discuss later in 
Section~\ref{s:design:method}. We  obtain a set of chains
of narrators that we feed to a morphological analyzer (MA) 
component controlled by a merger controller. 
The merger controller will look at the morphological analysis
of the narrator names and will decide to merge two narrators in 
one graph node in case they were morphologically equal. 
It will also give feedback to the MA component to stop the
analysis when it reaches a satisfactory result.
We consider this graph as a partial order relation between 
narrators where a narrator precedes another narrator if 
he narrated from him.

@@

The diagram in Figure~\ref{f:security} describes our 
high level vision for what the solution of the 
security case study
will look like in the CATARM framework. 
The inputs are two separate sets of documents. 
The first set includes crime scene investigation reports, 
and the second set includes 
interrogation reports with crime suspects as well as 
a database of previous interrogation reports. 
The CATARM system also takes as input a database of geographical
locations and a database of personal records. 
The aim is to check for inconsistencies in the interrogation
reports and to identify possible connections between the suspects
and the crime scene.

A morphological analyzer with date, place, and name extraction controller extracts
names, dates and places from the investigation reports. 
We pass the dates to a sorting component to produce $G_D$ 
a partial order graph between the dates. 
We look up the names in the personal records database and find out
$G_N$ a family dependency graph between the names. 
We pass the places to a mapper that builds $G_G$,
another graph relating the places using a geographical database.

We feed $G_D, G_F, G_G,$ and $G_R$   
to a consistency checker.
The consistency checker considers the three dimensions 
$N$ names, $G$ locations, and $D$ dates and projects $G_R$
over each one of them.
For each dimension $k$, it computes a local measure 
$\alpha_k$ of how much the projection $G_R|_k$ simulates $G_k$. 
It flags the components in the graph where the accumulated 
measure $\alpha= \Sigma_{k\in\{D,N,G\}} \alpha_k$ is
below a threshold $\theta$ as inconsistent.

@@

@@

Parental controls help parents allow their kids to benefit 
from the Internet 
without fearing the danger of violent and obscene material. 
These controls partially rely on processing the content to detect 
and rate the target material. 
Control mechanisms for Arabic traffic are currently mostly manual. 
In this case study we will detect the bad content. 
Given a set of undesired Internet content 
and a set of requests by Internet users, 
CATARM can discover whether there is a relation between 
the requested documents and the undesired content and flag the 
requested document with strong enough relation. 
The undesired content is available online and is in the 
order of thousands of documents. 
We will look at Arabic Internet traffic five times larger to
validate our case study.

@@

Clinical reports contain diagnosis information, observations about 
patient progress, prescriptions, and patient response to prescribed 
medications. 
In countries, like Lebanon, where medical doctors graduate from and 
specialize in 
several distinct practices around the world it is interesting to 
check for common practices and emerging trends to gain insight
about local and regional health and diagnosis features. 
It is also interesting to relate 
pharmaceutical 
records to clinical reports to check on 
market consistency and on self prescribed medicine 
practices.
Such a study can uncover serious tendencies in uncontrolled
pharmaceutical markets.
It can uncover what kind of sicknesses and what kind of
sensitive medicines are used without 
appropriate medical supervision.

@@

Morphological analyzers for the Arabic language
such as Buckwalter~\cite{Buckwalter:02},
Beesley~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} take an Arabic word as input
and produce several valid morphological solutions for the word. 
Each solution is a sequence of morphemes that form the word.
Each morpheme is associated with several tags such as 
morphological, part of speech (POS), grammar, and semantic tags. 
These analyzers are essential in several text analysis tools such 
as spell checkers as well as NLP frameworks~\cite{Col09}.
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy at the expense of efficiency.

@@

Case-based analysis can guide the morphological analysis
to prune lots of the solutions and to bias the order of considered
decisions. 
For example, in the hadith literature case study presented above, 
amongst other things, we are interested in checking whether a 
word is a name.  
We can bias the order of considered decisions to consider the 
morphemes that can be names first. 
If one of them matches, we can ignore the rest of the solutions. 
In case a prefix is decided to be of a morphological category
that does not connect with a name, then we can just ignore
the analysis of the rest of the word. 
This is more efficient than generating all possible partitions
of the word, checking for morphological correctness of each 
partition, and then inspecting the POS tags of each
correct solution. 
We desire to build a morphological analyzer that allows a 
case-based controller to intervene at every single decision
without much overhead.

Toutanova in~\cite{Tou09} suggested that using 
a joint model to do lemmatization and part of speech 
tagging performs better
than the direct and the pipelined approach. 
We take the first argument further to suggest a full joint NLP task 
model via using the NLP query to simplify and render the analysis
more accurate at the low level morphological analysis.

@@

We propose to build 
the CATARM query language with a user friendly
visual representation similar to that in Figures~\ref{f:literature} and
~\ref{f:security}. 
The basic elements in the query language will be entities, relations
and tags similar to those reported by a morphological analyzer. 
The query language will also have connectives such as set operations, 
relational operations, 
scalar and Boolean arithmetic operations, 
statistical computations, graph algorithms, and text algorithms
form terms and predicates that specify the desired computation
by the user. 
We will specify the CATARM query language in a formal grammar
that will be structured to resemble the question sentence
structure of the standard common Arabic.
This will be instrumental in identifying features
unique to the Arabic language in the analysis as we will
discuss later.

@@

We think that we can discover such features by building
a linguistic computational model (LCM) of the Arabic language
based on its formal grammar and formal morphology
as defined in several Arabic linguistic 
sources~\cite{Sha73,Abd00,Abd001}.
We will build our morphological analyzer on top 
of this LCM in addition to the current existing
concatenative-based analysis techniques.

Our LCM will differ from ElixirFM~\cite{Otakar:07}
in that it will follow a relational approach rather 
than a functional morphological approach. 
That is the solutions will be expressed in a directed acyclic graph 
(DAG)
representing a hierarchy of relations between the morphemes 
rather than morphological trees mapping the morphemes to the 
solutions.
We think this provides a more succinct
representation for the solutions.
Also interesting relations 
between morphemes can be computed statically before
the input is presented to the analyzer.
Morphological trees are traversals of these DAGs, are expensive 
to exhaustively compute and store statically, 
and are expensive to compute dynamically. 
Our intuition is that Arabic allows a higher order
of relational semantics in structures
such as noun sentences and complex hierarchical noun 
entities.

Let $L_q$ be the query language grammar, 
$L_A$ be the grammar that expresses the Arabic linguistic
computational model, and let $q$ be a user query expressed
in $L_q$.
We compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We will explore methods that use $\rho_q$ to simplify 
$L_A$ before we start the analysis. 
We think the possible simplifications that we can perform
here are Arabic language specific as they will reflect the 
correspondence between the formal query language definition $L_q$, 
the usage patterns of $L_q$  by Arabic users, 
and $L_A$ the formal LCM for Arabic.

@@

{\bf Case-based morphological analyzer.~~}
We experimented with several morphological analyzers.
We think that using finite state machines for morphological analysis
can be implemented in a more efficient way compared to other 
approaches. 
For example the concatenative rules used in Buckwalter~\cite{Tim04}
can be encoded in three FSMs with linear execution time using 
the trie~\cite{Aoe:89} data structure.

The use of FSMs also suits our case-based approach. 
The controller can be an FSM machine on its own and its decision
will be another input into the internal morphological FSM structures. 
This also suits our choice of enriching the concatenative analysis
with a formal linguistic model that can also be encoded into FSMs. 
FSM can also be easily integrated with hidden Markov models (HMM)
extracted from the sets of documents that the analyzer will consider.

@@

Our LCM will differ from ElixirFM~\cite{Otakar:07}
in that it will follow a relational approach.
In a relation, an element can have several associates.
In a function, an element is mapped to only one associate. 
We will express our morphological solutions in 
a directed acyclic graph (DAG)
representing a hierarchy of relations between the morphemes.
This is more succinct than morphological trees mapping the 
morphemes to the solutions as in~\cite{Otakar:07}.
Furthermore, interesting relations between the morphemes 
can be computed statically before
the input is presented to the analyzer.

In~\cite{Dao09}, Daoud observes that the pipeline model,
where morphology is performed first and syntactic and other
processes follow, is not adequate for Arabic. 
He suggests a synchronized syntactic and morphological
approach.  
We take the argument further to 
allow a tighter integration between the several pipeline
levels up to the NLP query in question by providing a 
case-based morphological analyzer, a linguistic computational
model of Arabic, and a query language that can all be expressed
as FSMs. 
The work~\cite{AlSham08} reports a similar result where
the automated addition of syntax knowledge improved stemming
in terms of accuracy and efficiency.

@@

{\bf Simplify LCM based on the query.~~}
Given $L_q$ the query language, 
$L_A$ the Arabic linguistic
computational model, and  $q$ a user query,
we compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We plan to inspect methods to over-approximate 
the morphemes and the transitions in $L_A$ 
that can be satisfied by morphemes and transitions
in $\rho_q$. 
This can simplify the number of states we need to 
look at under $L_A$.

@@

We will consider the use of HMM~\cite{JanHMM06}
models built on the fly while processing the documents.
A high level HMM model of the document builds an HMM state
model where the HMM states are morphemes or partial 
morphemes and the HMM edges represent transitions between
the HMM states. 
The HMM edges are labeled with weights that reflect the
probability of the transitions.
We will look at building a correspondence between 
the HMM model and the LCM and then simplify the 
LCM where the aggregated probabilities happen to be
below a certain threshold. 
The work in~\cite{JanHMM06} uses a similar approach
to correspond English text to tags corresponding to the 
Unified Medical Language System (UMLS).

@@

{\bf Entity and relational extractor.~~}
We will build a set of entity and relational extractors. 
The entities can be proper names, dates, locations,
and qualification adjectives. 
We will explore the use of pairing techniques to establish 
relations between the extracted entities.
The relations will form graphs as in the case studies presented
so far. The nodes represent the entities and the previously 
extracted relations and the edges represent the discovered 
relations.

@@

We plan to make use of existing annotated tree banks and
modern Arabic dictionaries
to augment the tags in the lexicons of the 
morphological analyzer and allow for better entity and relational
extractors.

{\bf Statistical components.~~}
We will make use and build standard and classical technical 
statistical components that can be used in CATARM. 
We will need statistical components to rank solutions, compute 
probable decisions, and estimate good thresholds. 
We can make use of current existing solutions such as the 
${\cal R}$ statistical library from the finance field.

@@

The solver passes the query to a front end that builds
a parse graph of the query. 
It uses the parse graph to simplify the LCM and the MA
as discussed before. 
Then it traverses the resulting parse graph. 
When the solver meets an atomic element of the query, 
it will call the morphological analyzer to compute 
morphological matches to the atomic element.
When the solver meets an operation of the query, 
it will call the corresponding computational component 
and passes to it the solutions computed so far from children
nodes. 
The solver will use existing off the shelf constraint solvers 
and set theory satisfiability solvers to address the queries. 
The solver may make use of existing 
propositional~\cite{MiniSAT04} or
higher order logic~\cite{Z308} solvers if needed. 
These solvers can answer satisfiability questions.

@@

We will follow a relaxed  eXtreme Programming 
methodology in the process of building CATARM.
This means that we will define refined incremental items of 
work that allow for more frequent deliverables. 
This also means that the code base will always be subject
to maintenance and refactoring to enhance reuse and 
increase productivity. 
While an item is delivered, it will still be subject to work,
enhancements, and maintenance. 
We selected eXtreme Programming as this fits well in the 
context of research where the clients are at close 
contact with the developers.
It accomodates the high dependency amongst the 
modules of CATARM. 
The pair programming practice in eXtreme Programming 
works well in the context of academy and research to keep
expertise in the team. 
As senior students graduate, their junior partners in the pair
will keep the knowledge in.

We detailed an aggressive work plan in Table~\ref{t:workplan}. 
The plan spans over three years, each with four quarters. 
The row headers describe work items, and the column headers
describe quarters in the corresponding year of the project.
The legends {\bf P, W, D, M, } 
and {\bf E} 
describe the type of activity taking place 
and stand for 
{\bf P:} {\em plan}, 
{\bf W:} {\em work }, 
{\bf D:} {\em deliver}, 
{\bf M:}  {\em maintain}, 
and 
{\bf E: } {\em enhance} respectively.
